# Lab 4: Analysis {.unnumbered}

You will start this session with your cleaned data ready to use in R. By the end of the session you will have computed the correlation statistic, produced some visualizations of your data, and be ready to present and write up your findings.


### Goals

- Understand what the correlation statistic quantifies
- Perform the appropriate correlational analysis on your data
- Interpret the results


## Correlation

Running with my example from last week, my variables were average extraversion scores and the Democratic Party feeling thermometer score. I made a data.frame with just those variables; filtered the data down to complete, valid responses; recoded the negatively-worded item; and computed an extraversion mean score. To refresh your memory, here's the entire pipeline from start to finish:


```{r}
#| warning: false
#| message: false

library(tidyverse)
library(anesr)
data(timeseries_2020)

my_data_complete <- timeseries_2020 |>
  select(trust = V201377, 
         concern = V201376) |> 
  haven::zap_labels() |>
  filter(if_all(everything(), ~ . >= 0)) |> 
  mutate(distrust = 6 - trust)

```


### The correlation statistic

The correlation statistic can be computed with a single line of code, as you'll see. But it's important to understand the math happening behind the scenes.



### Correlation in R



The data is ready to be analyzed. The correlation between two variables can be found using the `cor()` function.
```{r}

cor(x = my_data_complete$concern, 
    y = my_data_complete$distrust)

```


If you got an answer of `NA` instead of a number, it is probably because your data has some missing data. You just need to tell `cor()` to only use data for which both pairs of values are nonmissing:

```{r}

cor(x = my_data_complete$concern, 
    y = my_data_complete$distrust,
    use = "pairwise.complete.obs")

```

The `cor.test()` function goes further than `cor()`, giving you the $p$-value necessary for determining statistical significance^[Remember that, by convention, psychologists generally use $\alpha = .05$ as the criterion for statistical significance, meaning that if our data has less than a 5% chance of occurring under the null hypothesis we reject the null and tentatively accept the alternative hypothesis that the variables are associated.] and some other information about the correlation.

```{r}

cor.test(x = my_data_complete$concern, 
         y = my_data_complete$distrust)
```


Lastly, let's make a scatterplot visualizing the correlation.


```{r}
#| label: fig-scatter
#| fig-cap: Scatterplot (with jitter) of concern and distrust

my_data_complete |> 
  ggplot(aes(x = concern, y = distrust)) +
  # geom_point() +
  geom_point(position = position_jitter(width = 0.45, height = 0.45, seed = 1),
             alpha = 0.1) +
  scale_x_continuous(breaks = 1:5, 
                     labels = c(" Not at all", "A little", "Moderately", "Very", "Extremely")) +
  scale_y_continuous(breaks = 1:5,
                     labels = c("A great deal", "A lot", "A moderate amount", "A little", "None")) +
  labs(x = "Concern about government undermining the news media",
       y = "Trust in news media") +
  theme(panel.background = element_blank(),
        axis.line = element_line())

```


## Interpreting your findings

Remember, a correlation quantifies the general quantitative relationship between two sets of numbers; people's answers to your concern question are associated with their answer to the distrust question to the extent indicated by your correlation coefficient. To fully and fairly interpret this for your presentation and report you'll need to consider a number of things.

- What features of how the questions were asked might have affected people's answers?
- What, if anything, about the context in which the questions were asked might have influenced people's answers?
- How does it fit with the previous research you read about?

